KidsAIR initial PM data work

### 2020-01-10 dustrak_data_cleaning.Rmd
SUMs cleaning updates. Add extensive notes to PM and SUMs code.

---
title: "KidsAIR initial PM data work"
author: "Ethan Walker"
date: "Started 19 Nov 2019, Updated 20 Dec 2019"
output: pdf_document
---

@@

# Initial load, format, and save KidsAIR Dusttrak data
## Files here are in 3 formats: csv (2 different formats) and txt
## I had to do some manual cleaning of the files to get them named
## consistently and in reasonably similar formats to be able to load
## into R in batches, as seen below. The files I upload in the code below
## will also be uploaded to Box in a folder with my name.
## Box folder directory: 
```{r}
##### Read in dusttrack .txt files (from WMT) #####

# List files
list_txt_files <- list.files("Input/WMT/dusttrack_txt_files")

@@

### WMT ###
# List files
list_csv_files <- list.files("Input/WMT/dusttrack_csv_files")

@@

### NN ###
# List files
list_csv_files <- list.files("Input/NN/dusttrack_csv_files")

@@

### AK ###
# List files
list_csv_files <- list.files("Input/AK/dusttrack_csv_files")

@@

### WMT ### 
# List files
list_other_files <- list.files("Input/WMT/dusttrack_other_files")

@@

### NN ### 
# List files
list_other_files <- list.files("Input/NN/dusttrack_other_files")

@@

# Save pm dataset as RDS and CSV
write_rds(kids_pm, "Output/kids_pm_full_raw.rds")
#write_csv(kids_pm, "Output/kids_pm_full.csv")
```

@@

##### Add to new variables to Kids PM data #####
kids_pm_new <- kids_pm_joined %>%  
  #filter(area == "AK") %>% 
  # apply correction factor to pm and change to ug
  mutate(pm = pm/1.65,
         pm = pm*1000) %>% 
  mutate(home_winter_id = as.character(home_winter_id),
         area = as.factor(area),
         pm_date = date,
         pm_time = time,
         pm = as.numeric(pm),
         day_of_week = weekdays(date)) %>% 
  unite("pm_datetime", date:time, sep = " ") %>% 
  #select(home_winter_id, area, pm_date, pm_time, pm_datetime, day_of_week, pm) %>% 
  arrange(area, home_winter_id, pm_date) %>% 
  group_by(area, home_winter_id) %>% 
  # new variables summarizing pm2.5 within each home_winter_id
  mutate(pm_total_observations = n(),
         # new datetime values to calculate sampling interval
         min_datetime = first(pm_datetime),
         max_datetime = last(pm_datetime),
         min_datetime = ymd_hms(min_datetime),
         max_datetime = ymd_hms(max_datetime),
         diff_datetime = interval(min_datetime, max_datetime),
         # total sampling interval, based on first and last datetimes
         pm_sample_interval = diff_datetime/86400,
         # expected number of observations with the sampling interval
         expected_obs = pm_sample_interval*24*60,
         # difference between actual and expected observations
         actual_vs_expected_diff = pm_total_observations - expected_obs,
         # indicator variable if actual vs expected obs > 10 or greater
         actual_vs_expected = if_else(actual_vs_expected_diff < 10 &
                                      actual_vs_expected_diff > -10, 1, 0),
         pm_mean = mean(pm),
         # breaking sampling period into days
         obs = row_number(),
         sampling_day = if_else(obs < 1441, 1, 0),
         sampling_day = if_else(obs < 2881 & obs > 1440, 2, sampling_day),
         sampling_day = if_else(obs < 4321 & obs > 2880, 3, sampling_day),
         sampling_day = if_else(obs < 5761 & obs > 4320, 4, sampling_day),
         sampling_day = if_else(obs < 7201 & obs > 5760, 5, sampling_day),
         sampling_day = if_else(obs < 8641 & obs > 7200, 6, sampling_day),
         sampling_day = if_else(obs < 10081 & obs > 8640, 7, sampling_day),
         sampling_day = if_else(obs > 10080, 8, sampling_day)) %>%
  # calculate pm means for each sampling day
  group_by(area, home_winter_id, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm),
         pm_sampling_day = sampling_day) %>% 
  separate(starttime, c("trash1", "starttime"), sep = " ") %>% 
  separate(stoptime, c("trash2", "stoptime"), sep = " ") %>% 
  ungroup() %>% 
  select(-min_datetime, -max_datetime, -diff_datetime, -obs, -sampling_day, 
         -datetime, -trash1, -trash2)

# Save pm dataset as RDS and CSV
#write_rds(kids_pm_new, "Output/kids_pm_new.rds")
#write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

@@

# Run function from above to look at results
```{r}
# Load data
kids_pm <- read_rds("Output/kids_pm_new.rds")

@@

# Checking difference between dustrak start datetime and field data log datetime
sample_times <- kids_pm %>% 
  #filter(area == "AK") %>% 
  group_by(area, home_winter_id) %>% 
  arrange(pm_datetime) %>% 
  unite(start_datetime, c("startdate", "starttime"), 
        sep = " ", remove = FALSE) %>% 
  mutate(first_datetime = first(pm_datetime),
         first_datetime = ymd_hms(first_datetime),
         start_datetime = ymd_hms(start_datetime),
         time_diff = first_datetime - start_datetime) %>% 
  distinct(time_diff, .keep_all = TRUE) %>% 
  select(area, home_winter_id, pm_datetime, first_datetime,
         start_datetime, time_diff, startdate, starttime) %>% 
  ungroup() %>% 
  arrange(area, home_winter_id)
#write_csv(sample_times, "Output/pm_sample_times_comparison.csv")
```

# Loading and combining notes from time and PM logs
```{r}
time_notes <- read_xlsx("Input/pm_sample_times_comparison.xlsx") %>% 
  select(area, home_winter_id, first_datetime, start_datetime,
         time_diff, next_steps_time, check_log_time)

@@

notes_joined <- pm_notes %>% 
  left_join(time_notes, by = c("area", "home_winter_id")) %>% 
  left_join(kids_ids, by = c("area", "home_winter_id"))
#write_csv(notes_joined, "Output/pm_notes_check_logs.csv")
```

# Fix dusttrak sample start times
```{r}
# Load data
kids_pm <- read_rds("Output/kids_pm_new.rds")

@@

test <- kids_pm_new %>% 
  filter(percent_daily_obs_pm >= 0.9) %>% 
  summarise(n()/2241695)
# write_rds(kids_pm_new, "Output/kids_pm_new.rds")
# write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

>>>>

---
title: "KidsAIR initial PM data work"
author: "Ethan Walker"
date: "Started 19 Nov 2019, Updated 10 Jan 2020"
output: pdf_document
---

@@

# Initial load, format, and save KidsAIR DustTrak (PM) data
## Files were downloaded from Box: KidsAIR > 2019-Nov-Data > WMT > uploads > Dusttrack
## Files were not named consistently, so I renamed them by their home ID number.
## Files also had several different types and formats.
## Manual cleaning of some files had to be done to get them in 
## consistent formats and file types to work with in R.
## Unfortunately, this process had to be manual due to the large number of files
## and inconsistent QC process across the various files.
## As a result, this process will be difficult to replicate if needed;
## The code below will highlight how the files were formatted and grouped together.
## Once files were in consistent formats, I load them to R in batches, as seen below.
## The following code cleans and combines all PM data
```{r}
##### Read in dusttrack .txt files (from WMT) #####

# List files
## Numerous files had this format with a 28 row header
## 3 columns: date, time, PM
list_txt_files <- list.files("Input/WMT/dusttrack_txt_files")

@@

### WMT ###
# List files
## Numerous files had this format with no header
## 3 columns: date, time, PM
list_csv_files <- list.files("Input/WMT/dusttrack_csv_files")

@@

### NN ###
# List files
## Numerous files had this format with no header
## 3 columns: date, time, PM
list_csv_files <- list.files("Input/NN/dusttrack_csv_files")

@@

### AK ###
# List files
## Numerous files had this format with no header
## 3 columns: date, time, PM
list_csv_files <- list.files("Input/AK/dusttrack_csv_files")

@@

### WMT ### 
# List files
## 4 files had this format with a 19 row header
## 4 columns: elapsed time (start time extracted from header), PM, notes, notes
list_other_files <- list.files("Input/WMT/dusttrack_other_files")

@@

### NN ### 
# List files
## Numerous files had this format with a 19 row header
## 4 columns: elapsed time (start time extracted from header), PM, notes, notes
list_other_files <- list.files("Input/NN/dusttrack_other_files")

@@

# Save pm dataset as RDS and CSV
# write_rds(kids_pm, "Output/kids_pm_full_raw.rds")
# write_csv(kids_pm, "Output/kids_pm_full.csv")
```

@@

##### Add new variables to Kids PM data #####
kids_pm_new <- kids_pm_joined %>%  
  #filter(area == "AK") %>% 
  # apply correction factor to pm and change to ug
  mutate(pm = pm/1.65,
         pm = pm*1000) %>% 
  mutate(home_winter_id = as.character(home_winter_id),
         area = as.factor(area),
         pm_date = date,
         pm_time = time,
         pm = as.numeric(pm),
         day_of_week = weekdays(date)) %>% 
  unite("pm_datetime", date:time, sep = " ") %>% 
  #select(home_winter_id, area, pm_date, pm_time, pm_datetime, day_of_week, pm) %>% 
  arrange(area, home_winter_id, pm_date) %>% 
  group_by(area, home_winter_id) %>% 
  # new variables summarizing pm2.5 within each home_winter_id
  mutate(pm_total_observations = n(),
         # new datetime values to calculate sampling interval
         min_datetime = first(pm_datetime),
         max_datetime = last(pm_datetime),
         min_datetime = ymd_hms(min_datetime),
         max_datetime = ymd_hms(max_datetime),
         diff_datetime = interval(min_datetime, max_datetime),
         # total sampling interval, based on first and last datetimes
         pm_sample_interval = diff_datetime/86400,
         # expected number of observations with the sampling interval
         expected_obs = pm_sample_interval*24*60,
         # difference between actual and expected observations
         actual_vs_expected_diff = pm_total_observations - expected_obs,
         # indicator variable if actual vs expected obs > 10 or greater
         actual_vs_expected = if_else(actual_vs_expected_diff < 10 &
                                      actual_vs_expected_diff > -10, 1, 0),
         pm_mean = mean(pm),
         # breaking sampling period into days
         obs = row_number(),
         sampling_day = if_else(obs < 1441, 1, 0),
         sampling_day = if_else(obs < 2881 & obs > 1440, 2, sampling_day),
         sampling_day = if_else(obs < 4321 & obs > 2880, 3, sampling_day),
         sampling_day = if_else(obs < 5761 & obs > 4320, 4, sampling_day),
         sampling_day = if_else(obs < 7201 & obs > 5760, 5, sampling_day),
         sampling_day = if_else(obs < 8641 & obs > 7200, 6, sampling_day),
         sampling_day = if_else(obs < 10081 & obs > 8640, 7, sampling_day),
         sampling_day = if_else(obs > 10080, 8, sampling_day)) %>%
  # calculate pm means for each sampling day
  group_by(area, home_winter_id, sampling_day) %>% 
  mutate(pm_mean_daily = mean(pm),
         pm_sampling_day = sampling_day) %>% 
  separate(starttime, c("trash1", "starttime"), sep = " ") %>% 
  separate(stoptime, c("trash2", "stoptime"), sep = " ") %>% 
  ungroup() %>% 
  select(-min_datetime, -max_datetime, -diff_datetime, -obs, -sampling_day, 
         -datetime, -trash1, -trash2)

# Save pm dataset as RDS and CSV
# write_rds(kids_pm_new, "Output/kids_pm_new.rds")
# write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```

@@

# Run function from above to look at results
## Function run for each household to check for odd trends/gaps in the data
## Notes from this process on uploaded to Box:
### KidsAIR > Ethan - data cleaning and analysis > pm_cleaning_summary
```{r}
# Load data 
kids_pm <- read_rds("Output/kids_pm_new.rds")

@@

# Checking difference between dustrak start datetime and field data log datetime
sample_times <- kids_pm %>% 
  #filter(area == "AK") %>% 
  group_by(area, home_winter_id) %>% 
  arrange(pm_datetime) %>% 
  unite(start_datetime, c("startdate", "starttime"), 
        sep = " ", remove = FALSE) %>% 
  mutate(first_datetime = first(pm_datetime),
         first_datetime = ymd_hms(first_datetime),
         start_datetime = ymd_hms(start_datetime),
         time_diff = first_datetime - start_datetime) %>% 
  distinct(time_diff, .keep_all = TRUE) %>% 
  select(area, home_winter_id, pm_datetime, first_datetime,
         start_datetime, time_diff, startdate, starttime) %>% 
  ungroup() %>% 
  arrange(area, home_winter_id)
# write_csv(sample_times, "Output/pm_sample_times_comparison.csv")
```

# Loading and combining notes from time and PM logs
## Notes from the process of comparing actual vs expect start times 
## on uploaded to Box:
### KidsAIR > Ethan - data cleaning and analysis > pm_cleaning_summary
```{r}
time_notes <- read_xlsx("Input/pm_sample_times_comparison.xlsx") %>% 
  select(area, home_winter_id, first_datetime, start_datetime,
         time_diff, next_steps_time, check_log_time)

@@

notes_joined <- pm_notes %>% 
  left_join(time_notes, by = c("area", "home_winter_id")) %>% 
  left_join(kids_ids, by = c("area", "home_winter_id"))

# write_csv(notes_joined, "Output/pm_notes_check_logs.csv")
```

# Fix dusttrak sample start times
## Notes from above were used to make the following changes in the data
```{r}
# Load data 
kids_pm <- read_rds("Output/kids_pm_new.rds")

@@

test <- kids_pm_new %>% 
  filter(percent_daily_obs_pm >= 0.9) %>% 
  summarise(n()/2241695)

# write_rds(kids_pm_new, "Output/kids_pm_new.rds")
# write_csv(kids_pm_new, "Output/kids_pm_new.csv")
```
